---
output:
  pdf_document: default
  html_document: default
---
White wine properties and quality by M. Joukamaa
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(gridExtra)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
df <- read.csv('wineQualityWhites.csv')

# since the quality ratings given by the wine-tasting experts are discrete
# values on 0-10 scale, I am making a factored version of that variable in 
# order to facilitiate the use of the ratings in the analysis of the data.
df$quality.factored <- factor(df$quality)
```

In this report I am examining a dataset on white wine samples. The dataset
contains measurements about several chemical and physical qualities of the
samples as well as ratings given to them by wine-tasting experts.

# Univariate Plots Section

As a first step of the examination, I am creating univariate plots and numerical summaries of all the different variables present in the data. In these initial examinations I am considering  mostly the values from the lowest one of each
variable up to the 0.75 quantile of the value range. The reason for this is that
even the most cursory overview of the histograms and the value summaries below
shows that the maximum values are not particularly representative of the
distributions in most of the cases.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_fixed_acidity}
qplot(x = df$fixed.acidity, binwidth = 0.2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Summaries_fixed_acidity}
summary(df$fixed.acidity)
```

Fixed acidity ranges in 75 % of the samples from 3.8 grams per cubic decimetre
to 7.3 grams per cubic decimetre with a median of 6.8 grams per cubic decimetre.


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_vol_acidity}
qplot(x = df$volatile.acidity, binwidth = 0.02)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Summaries_vol_acidity}
summary(df$volatile.acidity)
```

Volatile acidity ranges from 0.08 grams per cubic decimeter to 0.32 grams per
cubic decimeter for 75 % of the samples with a median of 0.26 grams per cubic
decimetre.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_citric_acid}
qplot(x = df$citric.acid, binwidth = 0.02)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Summaries_citric_acid}
summary(df$citric.acid)
```

Citric acid concentrations range in 75 % of the samples from round zero to
0.39 grams per cubic decimetre with a median of 0.32 grams per cubic decimetre.
There is also a noticeable spike of samples for citric acid concentrations of
slightly below 0.5 grams per cubic decimetre.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_res_sugar}
qplot(x = df$residual.sugar, binwidth = 1)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Summaries_res_sugar}
summary(df$residual.sugar)
```

The quantities of residual sugar in the samples range from 0.6 grams per cubic
decimetre to 9.9 grams per cubic decimetre for 75 % of them with a median of 5.2
grams per cubic decimetre. Compared to the previous chemical and physical sample properties plotted, the distribution skews more notably to the right and is
especially concentrated on the lower end of the values.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_chlorides}
qplot(x = df$chlorides, binwidth = 0.005)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Summaries_chlorides}
summary(df$chlorides)
```

The amount of chlorides (essentially, salt) present ranges from 0.009 grams per
cubic decimeter to 0.05 grams per cubic decimeter for 75 % of the samples with a
median of 0.043 grams per cubic decimetre.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_free_SO2}
qplot(x = df$free.sulfur.dioxide, binwidth = 5)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Summaries_free_SO2}
summary(df$free.sulfur.dioxide)
```

The amount of free sulfur dioxide present ranges from 2 milligrams per cubic
decimetre to 46 milligrams per cubic decimetre for 75 % of the samples with a
median of 34 milligrams per cubic decimetre.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_total_SO2}
qplot(x = df$total.sulfur.dioxide, binwidth = 10)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Summaries_total_SO2}
summary(df$total.sulfur.dioxide)
```

The total amount of sulfur dioxide has a much wider range with the amount
varying between 9 milligrams per cubic decimetre and 167 milligrams per cubic
decimetre for 75 % of the samples with a median of 134 milligrams per cubic
decimetre.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_density}
qplot(x = df$density, binwidth = 0.001)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Summaries_density}
summary(df$density)
```

The density for most of the wine samples falls in a rather narrow range just
very slightly below 1 (that is, the density of water). This is understandable
given that wine contains ethanol, which has a density of only 0.789.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_pH}
qplot(x = df$pH, binwidth = 0.02)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Summaries_pH}
summary(df$pH)
```

For pH the range is also rather narrow one, varying from 2.72 to 3.28 for 75 %
of the samples with a median of 3.18. Of course, since the pH scale is
logarithmic, even these small differences can mean rather large and noticeable differences in the actual taste of the wine samples.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_sulphates}
qplot(x = df$sulphates, binwidth = 0.02)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Summaries_sulphates}
summary(df$sulphates)
```

Sulphates present in the samples range from 0.22 grams per cubic decimetre to
0.55 grams per cubic decimetre for 75 % of the samples with a median of 0.47
grams per cubic decimetre.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_alcohol}
qplot(x = df$alcohol, binwidth = 0.2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Summaries_alcohol}
summary(df$alcohol)
```

The proportion of alcohol (ethanol, to be precise) in the samples ranges from
8 % to 14.2 % of the sample volume with a median of 10.4 %. The histogram for
the variable shows that the distribution is somewhat right-tailed one, with a particularly large number of samples being in the 9.0 - 9.5 % alcohol volume
range. On the other hand, the alcohol content distribution doesn't have any real outliers, unlike the previous variables considered.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_quality}
qplot(x = df$quality.factored)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Summaries_quality}
summary(df$quality.factored)
```

Finally, there is the histogram and value summary for the wine quality samples
as judged by at least three wine-tasting experts. While the full rating scale
for quality is from 0 to 10, in the current dataset no sample gets a rating
lower than 3 or higher than 9. Majority of the ratings are concentrated in the
5-7 middle value range.

# Univariate Analysis

So, the dataset consists of 4898 observations of 13 different variables. Twelve
of these variables are actual measurements made on the various chemical/physical
qualities of the wine samples. The final variable is a rating on a 0-10 scale
given to the wine sample by at least three wine experts. As the expert rating
attempts to quantify the overall 'goodness' of a wine sample, this metric can be considered to be the response/dependent variable in relation to the chemical/
physical qualities. Also, while it is impossible for the expert ratings to
affect the measured qualities in any way, the measured qualities most probably
affect the expert ratings in some manner.

As the histograms above show, most of the chemical/physical qualities of the
wine samples tend to have a roughly normal distribution across the different
samples. However, the plots also show that pretty much all of the distributions
have more or less right-skewing tails. Related to this, it is also worth noting
that as the numerical summaries clearly show, the maximum values for almost all
of the above distributions are markedly larger than the upper bound of 0.75
quantile, which indicates that there are outliers present in the distributions.

There are also two variables which have distributions that don't really conform
to the properties of a normal distribution. As noted before, the amount of
residual sugar in the samples is heavily concentrated in the lowest levels of
the value range. Another variable that doesn't quite fit the normal distribution
mold is the alcohol content of the samples which is also somewhat concentrated
on the low end of the value range, though not as dramatically as the residual
sugar levels. I will apply some additional attention on these two variables in
the next section to make some more sense of these distributions.

The data seems tidy enough as it is, so any transformations or other operations
to tidy it up don't seem necessary. However, I did make a factor version of the
expert rating variable to be used alongside the original variable in those 
situations and examinations where it is necessary to use a factor-scale
variable.

# Bivariate Plots Section

Based on the considerations lined out above, the main question of interest
concerning the dataset is whether an association can be found between the chemical/physical qualities of a wine sample and the rating given to it by the wine-tasting experts. To investigate that question, I start by making boxplots
of all the said qualities grouped by the expert ratings as well as calculating
Pearson's correlation coefficients for these qualities in relation to the 
ratings.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_fixed_acidity}
ggplot(aes(x = quality.factored, y = fixed.acidity), data = df) +
  geom_boxplot(aes(group = quality))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlations_fixed_acidity}
cor.test(df$quality, df$fixed.acidity)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_vol_acidity}
ggplot(aes(x = quality, y = volatile.acidity), data = df) +
  geom_boxplot(aes(group = quality))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlations_vol_acidity}
cor.test(df$quality, df$volatile.acidity)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_citric_acid}
ggplot(aes(x = quality, y = citric.acid), data = df) +
  geom_boxplot(aes(group = quality))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlations_citric_acid}
cor.test(df$quality, df$citric.acid)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_res_sugar}
ggplot(aes(x = quality, y = residual.sugar), data = df) +
  geom_boxplot(aes(group = quality))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlations_res_sugar}
cor.test(df$quality, df$residual.sugar)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_chlorides}
ggplot(aes(x = quality, y = chlorides), data = df) +
  geom_boxplot(aes(group = quality))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlations_chlorides}
cor.test(df$quality, df$chlorides)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_free_SO2}
ggplot(aes(x = quality, y = free.sulfur.dioxide), data = df) +
  geom_boxplot(aes(group = quality))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlations_free_SO2}
cor.test(df$quality, df$free.sulfur.dioxide)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_total_SO2}
ggplot(aes(x = quality, y = total.sulfur.dioxide), data = df) +
  geom_boxplot(aes(group = quality))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlations_total_SO2}
cor.test(df$quality, df$total.sulfur.dioxide)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_density}
ggplot(aes(x = quality, y = density), data = df) +
  geom_boxplot(aes(group = quality))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlations_density}
cor.test(df$quality, df$density)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_pH}
ggplot(aes(x = quality, y = pH), data = df) +
  geom_boxplot(aes(group = quality))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlations_pH}
cor.test(df$quality, df$pH)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_sulphates}
ggplot(aes(x = quality, y = sulphates), data = df) +
  geom_boxplot(aes(group = quality))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlations_sulphates}
cor.test(df$quality, df$sulphates)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_alcohol}
ggplot(aes(x = quality, y = alcohol), data = df) +
  geom_boxplot(aes(group = quality))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlations_alcohol}
cor.test(df$quality, df$alcohol)
```

I am also examining if there are observeable associations between some of the
other variables of the dataset as well. I am first taking a look at the
relationship between the alcohol content of a wine sample and the amount of
residual sugar present in it. As the sugar of the grapes tranforms into ethanol
during the fermentation process of the wine, my initial assumption would be that
those wine samples with a higher alcohol content would have less residual sugar
present in them.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_alcohol_res_sugar}
ggplot(aes(x = alcohol, y = residual.sugar), data = df) +
  geom_point(alpha = 0.05, position = 'jitter')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlations_alcohol_res_sugar}
cor.test(df$alcohol, df$residual.sugar)
```

I am also interested to see if those samples with a higher alcohol content have
a lower density than those with less alcohol in them, as the density of ethanol
is less than that of water.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_alcohol_density}
ggplot(aes(x = alcohol, y = density), data = df) +
  geom_point(alpha = 0.1, position = 'jitter')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlations_alcohol_density}
cor.test(df$alcohol, df$density)
```

# Bivariate Analysis

For most of the chemical/physical qualities of the wine samples there doesn't
seem to be any particularly noteworthy variation when they are examined across
the expert ratings. The value ranges for most of the variables are rather
narrow to begin with, and especially when looking at the median values, the 
variation doesn't seem to exhibit discernible patterns for most of the
variables. Also, most of the coefficients calculated for correlations between
the quality ratings and other variables are not particularly strong.

However, a few of the chemical/physical qualities of the samples exhibit an
association with the expert ratings that are clear and consistent enough to
warrant further examination. In terms of the correlation coefficients, the
strongest associations that the expert ratings have are with alcohol content
(r = 0.44), density (r = -0.31), chlorides (r = -0.21), volatile acidity
(r = -0.19), and total amount of sulfur dioxide (r = -0.17).

When looking at the boxplots of these variables by the levels of expert ratings,
they mostly show similar relationships and trends as indicated by the
correlation coefficients. However, it is worth noting that the boxplots for
chloride and volatile acidity levels show numerous instances of outliers in the
distributions of these variables. This factor calls for prudence in further
analyses concerning these two variables.

With regards to the density of the samples, the scatter plot of sample density
vs. alcohol content which ends the previous section shows a clear (and linear)
relationship between these two variables. This is also affirmed by the strong
correlation coefficient (r = -0.78) of the variables. Thus, the density of
the samples can be considered to be mostly a function of their alcohol content
and should be dropped from further investigations as a superfluous and possibly
confounding variable.

However, given that the alcohol content of the samples affects their density in
observeable amounts, the couple of outliers that have a density above that of
water invite some interesting speculation. Could it be that the samples in
question have also some foreign substances present in them that increase their
density to a marked degree?

Finally, the plot of residual sugar vs. alcohol content presents a more complex relationship. While in general terms the samples with higher alcohol levels tend
to have lower amounts of residual sugar, there are also samples with low levels
of both alcohol and residual sugar. So while there are wine samples where the fermentation process has not processed all of the sugar into ethanol, there are
probably also grape strains with lower levels of sugar to begin with - and thus
also low levels of alcohol in the end product as well.

# Multivariate Plots Section

In the previous section four variables were identified which exhibited
noteworthy relationships with the quality ratings of the samples: alcohol
content, total amount of sulfur dioxide, volatile acidity and chloride levels of
the samples. Next, I am making several scatter plots to see if the pairwise
plotting of these chemical/physical qualities grouped by the expert rating
levels shows any clear trivariate relationships between these variables.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_volatile_acidity_vs_chlorides}
ggplot(aes(x = volatile.acidity, y = chlorides), data = df) +
  geom_point(aes(color = quality.factored), alpha = 0.3, position = 'jitter') +
  scale_color_brewer(type = 'div', direction = -1)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_volatile_acidity_vs_total_SO2}
ggplot(aes(x = volatile.acidity, y = total.sulfur.dioxide), data = df) +
  geom_point(aes(color = quality.factored), alpha = 0.3, position = 'jitter') +
  scale_color_brewer(type = 'div', direction = -1)
```

```{r echo=FALSE, message=FALSE, warning=FALSE,  Multivariate_Plots_alcohol_vs_chlorides}
ggplot(aes(x = alcohol, y = chlorides), data = df) +
  geom_point(aes(color = quality.factored), alpha = 0.3, position = 'jitter') +
  scale_color_brewer(type = 'div', direction = -1)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_alcohol_vs_volatile_acidity}
ggplot(aes(x = alcohol, y = volatile.acidity), data = df) +
  geom_point(aes(color = quality.factored), alpha = 0.3, position = 'jitter') +
  scale_color_brewer(type = 'div', direction = -1)
```

```{r echo=FALSE, message=FALSE, warning=FALSE,     Multivariate_Plots_total_SO2_vs_chlorides}
ggplot(aes(x = total.sulfur.dioxide, y = chlorides), data = df) +
  geom_point(aes(color = quality.factored), alpha = 0.3, position = 'jitter') +
  scale_color_brewer(type = 'div', direction = -1)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_alcohol_vs_total_sulfur_dioxide}
ggplot(aes(x = alcohol, y = total.sulfur.dioxide), data = df) +
  geom_point(aes(color = quality.factored), alpha = 0.3, position = 'jitter') +
  scale_color_brewer(type = 'div', direction = -1)
```

# Multivariate Analysis

Most of the scatter plots I made above don't really present any discernible
patterns of trivariate relationships between the two variables plotted and the
grouping made by expert ratings. While in some of the plots the best-rated wines
are clearly concentrated in a certain end of one of the axes, the expert ratings
tend to be all over the place in terms of the other plotted variable. The
numerous outliers present in the volatile acidity and chloride distributions
that were noted previously account undoubtedly for a lot of this noise. Still,
the lack of discernible patterns in most of the above plots doesn't probably
return exclusively to the effect of the outliers.

However, the scatter plot of total sulfur dioxide amount versus alcohol content
seems to show a clear and discernible relationship between the three different
variables examined. The best-rated wines are concentrated mostly not only in the
high end of the alcohol volume scale but also in the lower ratings of total
sulfur dioxide amount. Thus, according to these results, in order for a wine to
achieve a good expert rating it needs to have a relatively high alcohol volume,
but the amount of sulfur dioxide in it should also be rather low.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
ggplot(aes(x = quality.factored, y = alcohol), data = df) +
  geom_boxplot(aes(group = quality.factored), fill = '#DDEEEE') +
  labs(x = 'Quality rating (0-10)', y = 'Alcohol (vol. %)',
       title = 'Alcohol content by expert quality ratings') +
  theme_light()
```

### Description One
The plot above shows a strong and clear relationship between the alcohol
content of the wine samples and expert ratings given to them. This relationship
is not entirely straightforward, though. As can be seen in the plot, there is a
slight upward trend in the alcohol content of the samples toward the lower range
of the ratings scale as well. But in general it can be said that wine samples
with higher alcohol content tend to also get higher expert ratings.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(aes(x = quality.factored, y = total.sulfur.dioxide), data = df) +
  geom_boxplot(aes(group = quality.factored), fill = '#DDEEEE') +
    labs(x = 'Quality rating (0-10)', y = 'Total sulfur dioxide (mg/dm³)',
         title = 'Total sulfur dioxide by expert quality ratings') +
  theme_light()
```

### Description Two
This second plot show that for the total amount of sulfur dioxide in the samples
the higher ratings are associated with a downward trend for the sulfur dioxide.
Even though there can be seen some slight fluctuation in the plot, the overall
trend is quite clear.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(x = alcohol, y = total.sulfur.dioxide), data = df) +
  geom_point(aes(color = quality.factored), alpha = 0.3, position = 'jitter') +
  scale_y_sqrt() +
  scale_color_brewer(type = 'div', direction = -1) +
  labs(color = 'Quality rating', x = 'Alcohol (vol. %)',
       y = 'Total sulfur dioxide (mg/dm³), square root scale',
       title = 'Alcohol content vs. total sulfur dioxide',
       subtitle = 'Grouped by expert quality ratings') +
  theme_light()
```

### Description Three
Combining the previous two examinations, the above plot shows how the color
gradient for expert ratings moves quite clearly along both axes of the plot.
Thus the best-rated wines are mostly in the high end of the alcohol content
scale and low end of the sulfur dioxide amount scale with both of the variables
having a roughly linear relationship with the expert ratings. When appraising
the plot, it should be noted that the y-axis of the plot has been transformed
into a square root scale in order to make the plot clearer and more readable.

------

# Reflection

In this report I set out to examine a dataset of 4898 wine samples with 13 
different variables measuring the chemical and physical qualities of the samples
as well as rating them on their overall quality. I looked over the distributions
of the individual variables and then focused more closely on how the different
chemical and physical qualities were associated with quality ratings of the
samples. In the end, I found a rather clear and straightforward three-way
relationship between the sample ratings, their alcohol content and the total
amount of sulfur dioxide in them.

It was somewhat sobering (pun intended) to see how strongly the alcohol content
was correlated with the ratings given to the samples by wine-tasting experts, as
even these experts seem to mostly favor strong wines over lighter ones. While
some of the other objective properties of the wine samples were also associated
with the expert rating levels to some degree, the negative relationship with the
total amount of sulfur dioxide provided a particularly distinguishable fit when
coupled with the alcohol content.

The examination of the dataset was mostly a rather straightforward job, though
of course time-consuming in nature. Probably the most vexing part of the
examination was figuring out which of the bivariate relationships were
significant enough to warrant further investigation. Stylistical considerations
for the three summary plots necessitated also several trial-and-error iterations
before arriving to their final designs.

While further analyses about the relationships between the different chemical
and physical properties of the samples can always be made, more fruitful work on
the samples would probably benefit from additional data. For example, 
information about grape strains used or fermentation times and conditions could
probably provide all kinds of interesting insights when combined with the 
current data. Another possible avenue of investigation would be to look at the
prices of the different wines and how much they correlate on one hand with the
expert ratings, and on the other hand with the most significant chemical and
physical properties of the samples.